---
title: "Week 6"
author: "Reesa Devers"
date: "2025-02-19"
output: html_document
---

```{r}
setwd("/Users/reesadevers/Desktop/BIOSTAT/")
cod.df <- read.csv(file= "Cod.csv")
vec <- c("Reesa's", "BIO 380", "Lab")

my.vec <- seq(from = 10, to = 300, length = 5)

another.vec <- rep(c("Success", "Failure"), times = c(30, 70))
vec[2]
my.vec[3]
another.vec[100]
another.vec[50:60]
vec[-2]
vec[-c(2:3)]

a.matrix <- matrix(data = LETTERS[1:20], ncol = 4, nrow = 5, byrow = T)
a.matrix[2, 3]
a.matrix[2, ]
a.matrix[, 3]
a.matrix[-3, c(3:4)]
```

```{r}
str(cod.df)
cod.df$Prevalence
```

```{r}
cod.df[cod.df$Sex == 2, "Sex"] <- "Male"
cod.df[cod.df$Sex == 1, "Sex"] <- "Female"
cod.df[cod.df$Sex == 0, "Sex"] <- "Unknown"
head(cod.df$Sex)
```

```{r}
Male1.df <- subset(cod.df, subset = (Sex == "Male"))
Male2.df <- cod.df[cod.df$Sex == "Male", ]
identical(Male1.df, Male2.df)
```
```{r}
clean.cod.df <-subset(cod.df, subset = (Sex != "Unknown"))
str(clean.cod.df)

dim(clean.cod.df)

unique(clean.cod.df$Sex)

table(clean.cod.df$Sex)

table(clean.cod.df$Sex, clean.cod.df$Stage)

aggregate(Length ~ Sex, data = clean.cod.df, sd)
```

#Checkpoint 1: clean.cod.df <-subset(cod.df, subset = (Sex != "Unknown"))

```{r}
table(cod.df$Sex)
total <- 1254
Female <- 574
Male <- 598
Unknown <- 82

Female/total
Male/total
Unknown/total

table(cod.df$Area)

aggregate(Weight ~ Sex, data = cod.df, mean)

parasite_table <- table(cod.df$Age,cod.df$Prevalence, cod.df$Sex)

parasite_prevalence <- prop.table(parasite_table, margin = c(1,2))



```
```{r}

parasites_table <- table(cod.df$Area,cod.df$Prevalence)

parasite_prevalences <- prop.table(parasites_table, margin = c(1,2))


```

#Checkpoint 2: 
#A.table(cod.df$Sex).   Number of fish = 1254

#B.total <- 1254
#Female <- 574
#Male <- 598
#Unknown <- 82
#Female/total
#Male/total
#Unknown/total

#C.table(cod.df$Area)   4 Areas Total

#D.aggregate(Weight ~ Sex, data = clean.cod.df, mean)

#E.Infection is more common in Males, code below:
#parasite_table <- table(cod.df$Age,cod.df$Prevalence, cod.df$Sex)
#parasite_prevalence <- prop.table(parasite_table, margin = c(1,2))

#F. Infection more common in areas 3 and 4. Code below:
#parasites_table <- table(cod.df$Area,cod.df$Prevalence)
#parasite_prevalences <- prop.table(parasites_table, margin = c(1,2))


```{r}
barplot(table(cod.df$Sex), xlab = "Sex", ylab = "Number of Fish")
mosaicplot(Sex ~ 1, data = cod.df, main = "Sex Distribution in Cod")
barplot(table(cod.df$Sex), xlab = "Sex", ylab = "Number of Fish", col = "red")
install.packages("viridisLite")

library(viridisLite)
```

```{r}
(our.colors <- mako(length(unique(cod.df$Sex)), begin = 0.25, end = 0.75))
barplot(table(cod.df$Sex), xlab = "Sex", ylab = "Number of Fish", col = our.colors,
main = "Sex Distribution of Norwegian Cod", cex.lab = 1.5)

par(mfrow = c(2, 2))
boxplot(cod.df$Weight)
plot(density(cod.df$Weight, na.rm = TRUE))
hist(cod.df$Weight)
stripchart(cod.df$Weight, method = "jitter", pch = 19)

par(mfrow = c(2, 2))
boxplot(cod.df$Weight, horizontal = TRUE)
plot(density(cod.df$Weight, na.rm = TRUE))
hist(cod.df$Weight)
stripchart(cod.df$Weight, method = "jitter", pch = 19, col = rgb(0, 0, 0, alpha = 0.05))

summary(cod.df$Weight)


```

```{r}
par(mfrow = c(1, 2))
plot(cod.df$Area, cod.df$Weight, xlab = "Area", ylab = "Fish Weight", pch = 19, col = rgb(0,
0, 0, alpha = 0.5), main = "Without Jitter")
plot(jitter(cod.df$Area), cod.df$Weight, xlab = "Area", ylab = "Fish Weight", pch = 19,
col = rgb(0, 0, 0, alpha = 0.05), main = "With Jitter")

head(cod.df$Area)
head(jitter(cod.df$Area))
plot(cod.df$Length, cod.df$Weight, pch = 19, xlab = "Length", ylab = "Weight", col = rgb(0,
0, 0, alpha = 0.15))
my.colors <- viridis(2, begin = 0.3, end = 0.7, alpha = 0.25)
male.df <- subset(cod.df, subset = (Sex == "Male"))
female.df <- subset(cod.df, subset = (Sex == "Female"))
plot(female.df$Length, female.df$Weight, pch = 19, cex = 1.5, xlab = 'Length', ylab = 'Weight', col = my.colors[1])
points(male.df$Length, male.df$Weight, pch = 19, cex = 1.5, col = my.colors[2])
legend('topleft', #position of legend
legend = c("Females","Males"), col = my.colors, pch = 19, cex = 1.5, bty = 'n') 

pdf(file = 'MyFishPlot.pdf')

plot(female.df$Length, female.df$Weight,
pch = 19, cex = 1.5, xlab = 'Length', ylab = 'Weight', col = my.colors[1])
points(male.df$Length, male.df$Weight, pch = 19, cex = 1.5, col = my.colors[2])
legend('topleft',
legend = c("Females","Males"),  col = my.colors,pch = 19, cex = 1.5,  bty = 'n') 
dev.off()
```
#Checkpoint 3:
#barplot(table(cod.df$Sex, cod.df$Prevalence), xlab = "Prevalence", ylab = "Sex", col = our.colors, main = "Prevalence of disease in different Sex's of Cod", cex.lab = 1.5)

```{r}
barplot(table(cod.df$Sex, cod.df$Prevalence), xlab = "Prevalence", ylab = "Sex", col = our.colors, main = "Prevalence of disease in different Sex's of Cod", cex.lab = 1.5)

malef.df <- subset(cod.df, subset = (Sex == "Male"))
femalef.df <- subset(cod.df, subset = (Sex == "Female"))

plot(femalef.df$Prevalence, femalef.df$Area,
pch = 19, cex = 1.5,
xlab = 'Prevalence', ylab = 'Area',
col = "red")

points(malef.df$Prevalence, malef.df$Area,
pch = 19, cex = 1.5,
col = "blue")

plot(cod.df$Prevalence ~ cod.df$Area, pch = 19, xlab = "Area", ylab = "Prevalence", col = our.colors, main = "Prevalence of disease in different Sex's of Cod")

```

#Checkpoint 4:
#plot(cod.df$Prevalence ~ cod.df$Area, pch = 19, xlab = "Area", ylab = "Prevalence", col = our.colors, main = "Prevalence of disease in different Sex's of Cod")

```{r}
plot(jitter(cod.df$Area), cod.df$Prevalence, xlab = "Area", ylab = "Prevalence", pch = 19, col = rgb(0, 0, 0, alpha = 0.05), main = "With Jitter")
```



#Checkpoint 5:plot(jitter(cod.df$Area), cod.df$Prevalence, xlab = "Area", ylab = "Prevalence", pch = 19, col = rgb(0, 0, 0, alpha = 0.05), main = "With Jitter")


```{r}
alphabet <- letters[1:26]
sample(alphabet, size = 10, replace = T)
sample(alphabet, size = 10, replace = F) 
rbinom(5, size = 10, prob = 0.6)
dbinom(4:7, size = 10, prob = 0.6)
pbinom(4:7, size = 10, prob = 0.6)
qbinom(c(0.5, 0.9), size = 10, prob = 0.6)
sample.size <- 10
prob.success <- 0.6
possible.outcomes <- 0:sample.size
probabilities <- dbinom(possible.outcomes, size = sample.size, prob = prob.success)
par(mfcol = c(2, 2), mar = c(4, 4, 3, 1))
plot(possible.outcomes, probabilities, xlab = "# Successes", ylab = "Probability", main = "Probability Distribution Function", pch = 19)
cumulative.prob <- pbinom(possible.outcomes, size = sample.size, prob = prob.success)
plot(possible.outcomes, cumulative.prob, xlab = "# Successes", ylab = "Cumulative Probability",main = "Cumulative Distribution Function", pch = 19)
hist(rbinom(100, size = sample.size, prob = prob.success), xlab = "# Successes",
main = "Random Sample")
quantile.probs <- seq(from = 0, to = 1, length = 100)
quantiles <- qbinom(quantile.probs, size = sample.size, prob = prob.success)
plot(quantile.probs, quantiles, xlab = "Cumulative Probability", ylab = "Quantile", main = "Quantile Function", typ = "l")

mu <- 10
sigma <- 2
n <- 30
xbar.mu <- mu
xbar.sd <- sigma/sqrt(n)

pot.outcomes <- seq(from = 5, to = 15, length = 10000)
sam.prob <- dnorm(pot.outcomes, mean = xbar.mu, sd = xbar.sd)
sam.cumul.prob <- pnorm(pot.outcomes, mean = xbar.mu, sd = xbar.sd)
par(mfcol = c(1, 2), cex.lab = 0.8, cex.axis = 0.8)
plot(pot.outcomes, sam.prob, typ = "l", xlab = "Sample Outcome", ylab = "Probability",
main = "Sampling Distribution")
plot(pot.outcomes, sam.cumul.prob, typ = "l", xlab = "Sample Mean", ylab = "Cumulative Probability",
main = "Sampling Distribution")
```
```{r}
sample_size <- 5

sample.df <- sample_size - 1  
x <- seq(-5, 5, by = 0.01)
y <- dt(x, df = sample.df) 

plot(x, y, type = "l", xlab = "t", ylab = "Density", main = "t-distribution (df = 4)")

lower_quantile <- qt(0.025, df = sample.df) 

upper_quantile <- qt(0.975, df = sample.df)
abline(v = lower_quantile, col = 'red', lty = 2)
abline(v = upper_quantile, col = 'red', lty = 2)
```

#Checkpoint 6: sample_size <- 5
#sample.df <- sample_size - 1  
#x <- seq(-5, 5, by = 0.01)
#y <- dt(x, df = sample.df) 
#plot(x, y, type = "l", xlab = "t", ylab = "Density", main = "t-distribution")

#lower_quantile <- qt(0.025, df = sample.df) 
#upper_quantile <- qt(0.975, df = sample.df)
#abline(v = lower_quantile, col = 'red', lty = 2)
#abline(v = upper_quantile, col = 'red', lty = 2)


```{r}
sample_size <- 50

sample.df <- sample_size - 1  
x <- seq(-5, 5, by = 0.01)
y <- dt(x, df = sample.df) 

plot(x, y, type = "l", xlab = "t", ylab = "Density", main = "t-distribution")

lower_quantile <- qt(0.025, df = sample.df) 

upper_quantile <- qt(0.975, df = sample.df)
abline(v = lower_quantile, col = 'red', lty = 2)
abline(v = upper_quantile, col = 'red', lty = 2)
```

#Checkpoint 7:
#sample_size <- 50
#sample.df <- sample_size - 1  
#x <- seq(-5, 5, by = 0.01)
#y <- dt(x, df = sample.df) 
#plot(x, y, type = "l", xlab = "t", ylab = "Density", main = "t-distribution")
#lower_quantile <- qt(0.025, df = sample.df) 
#upper_quantile <- qt(0.975, df = sample.df)
#abline(v = lower_quantile, col = 'red', lty = 2)
#abline(v = upper_quantile, col = 'red', lty = 2)

```{r}
n <- length(cod.df$Prevalence)
num.infected <- sum(cod.df$Prevalence == 1)
alpha <- 0.01
conf.levels <- c(alpha/2, 1 - alpha/2)
(conf.int <- qbinom(conf.levels, size = n, prob = num.infected/n))/n
```
#Checkpoint 8:n <- length(cod.df$Prevalence)
#num.infected <- sum(cod.df$Prevalence == 1)
#alpha <- 0.01
#conf.levels <- c(alpha/2, 1 - alpha/2)
#(conf.int <- qbinom(conf.levels, size = n, prob = num.infected/n))/n

```{r}
hist(cod.df$Length)
mean.length <- mean(cod.df$Length, na.rm = T)
n.length <- sum(!is.na(cod.df$Length))
sd.length <- sd(cod.df$Length, na.rm = T)
se.length <- sd.length/sqrt(n.length)
n.length
mean.length
sd.length
se.length
```
```{r}
alpha <- 0.01
conf.levels <- c(alpha/2, 1 - alpha/2)
qt(conf.levels, df = n.length - 1)

(fish.conf.int <- mean.length + qt(conf.levels, df = n.length - 1) * se.length)

length.mdl <- lm(Length ~ 1, data = cod.df)
confint(length.mdl, level = 0.99)
```
```{r}

length.mdl <- lm(Length ~ 1, data = cod.df)
confint(length.mdl, level = 0.95)
```

#Checkpoint 10:
#length.mdl <- lm(Length ~ 1, data = cod.df)
#confint(length.mdl, level = 0.95)



```{r}

```


#Checkpoint 11:





