---
title: "Week 5"
author: "Reesa Devers"
date: "2025-02-12"
output: html_document
---

```{r}
setwd("/Users/reesadevers/Desktop/BIOSTAT/")
diabetes.df <- read.csv(file= "diabetes.csv")
num.frogs.sampled <- 5
prob.inf <- 0.6
(poss.outcomes <- 0:num.frogs.sampled)

(outcomes.probs <- dbinom(poss.outcomes, size = num.frogs.sampled, p = prob.inf))

plot(poss.outcomes, outcomes.probs, pch = 19,
xlab = '# Infected Frogs in Sample', ylab = 'Probability',
main = 'Sampling Distribution for Number of Infected Frogs')

num.frogs.sampled <- 30
prob.inf <- 0.6
poss.outcomes <- 0:num.frogs.sampled
outcomes.probs <- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf)
inf.prevalence <- poss.outcomes/num.frogs.sampled
plot(inf.prevalence, outcomes.probs, pch = 19,
xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
```

```{r}
num.frogs.sampled <- 30
prob.inf <- 0.6
poss.outcomes <- 0:num.frogs.sampled
outcomes.probs <- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf)
inf.prevalence <- poss.outcomes/num.frogs.sampled
plot(inf.prevalence, outcomes.probs, pch = 19,
xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
```

#Checkpoint 1: 40% of frogs infected is the bounds for reasonable sample prevalence in this case.

\############# NK Comments \##################

You only have a single number. I was expecting two: one lower bound and one upper bound. Stated differently, I was asking what you might tell someone you expect to see in their sample of frogs.

\############# NK Comments \##################

```{r}
cumul.prob <- pbinom(poss.outcomes, size = num.frogs.sampled, p = prob.inf)
plot(inf.prevalence, cumul.prob, pch = 19, xlab = 'Infection Prevalence in a Sample',
ylab = 'Cumulative Probability')

qbinom(0.25, size = num.frogs.sampled, prob = prob.inf)

qbinom(0.75, size = num.frogs.sampled, prob = prob.inf)

plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
abline(v = c(16,20)/num.frogs.sampled, lty = 2, col = 'red')

qbinom(0.025, size = num.frogs.sampled, prob = prob.inf)

qbinom(0.975, size = num.frogs.sampled, prob = prob.inf)

plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
abline(v = c(13,23)/num.frogs.sampled, lty = 2, col = 'red')


```

```{r}
num.frogs.sampled <- 100
prob.inf <- 0.6
poss.outcomes <- 0:num.frogs.sampled
outcomes.probs <- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf)
inf.prevalence <- poss.outcomes/num.frogs.sampled
plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')

qbinom(0.2, size = num.frogs.sampled, prob = prob.inf)

qbinom(0.8, size = num.frogs.sampled, prob = prob.inf)

plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
abline(v = c(56,64)/num.frogs.sampled, lty = 2, col = 'red')

```

#Checkpoint 2: #num.frogs.sampled \<- 100 #prob.inf \<- 0.6 #poss.outcomes \<- 0:num.frogs.sampled #outcomes.probs \<- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf) #inf.prevalence \<- poss.outcomes/num.frogs.sampled #plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '\# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')

#qbinom(0.025, size = num.frogs.sampled, prob = prob.inf)

#qbinom(0.975, size = num.frogs.sampled, prob = prob.inf)

#plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '\# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence') #abline(v = c(50,69)/num.frogs.sampled, lty = 2, col = 'red')

\############# NK Comments \##################

Looks good

\############# NK Comments \##################

```{r}
num.frogs.sampled <- 100
prob.inf <- 0.6
poss.outcomes <- 0:num.frogs.sampled
outcomes.probs <- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf)
inf.prevalence <- poss.outcomes/num.frogs.sampled
plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')

qbinom(0.025, size = num.frogs.sampled, prob = prob.inf)

qbinom(0.975, size = num.frogs.sampled, prob = prob.inf)

plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')
abline(v = c(50,69)/num.frogs.sampled, lty = 2, col = 'red')
```

#Checkpoint 3: #num.frogs.sampled \<- 100 #prob.inf \<- 0.6 #poss.outcomes \<- 0:num.frogs.sampled #outcomes.probs \<- dbinom(poss.outcomes, size = num.frogs.sampled, prob = prob.inf) #inf.prevalence \<- poss.outcomes/num.frogs.sampled #plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '\# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence')

#qbinom(0.025, size = num.frogs.sampled, prob = prob.inf)

#qbinom(0.975, size = num.frogs.sampled, prob = prob.inf)

#plot(inf.prevalence, outcomes.probs, pch = 19, xlab = '\# Infected Frogs in Sample', ylab = 'Probability', main = 'Sampling Distribution for Infection Prevalence') #abline(v = c(50,69)/num.frogs.sampled, lty = 2, col = 'red')

\############# NK Comments \##################

Looks good.

\############# NK Comments \##################

#Checkpoint 4: Because we had more individuals sampled in this sample of 100 compared to 10, our distribution is even more closely related to a normal distribution. There is less variance in how the tail ends of the distribution look with more individuals. The graph is more semetrical and the mean is the same: 3.

\############# NK Comments \##################

One of the most important differences is the narrower range of possible values. Compare the x-axes of the two graphs. You can see that the sample mean is much less variable sample to sample when there are 100 individuals in the sample.

\############# NK Comments \##################

```{r}
fish.size.sd <- 10
sample.size <- 20
SE.mean.fish.size <- fish.size.sd/sqrt(sample.size)
SE.mean.fish.size
```

```{r}
fish.size.sd <- 10
sample.size <- 30
SE.mean.fish.size <- fish.size.sd/sqrt(sample.size)
SE.mean.fish.size

fish.size.sd <- 10
sample.size <- 50
SE.mean.fish.size <- fish.size.sd/sqrt(sample.size)
SE.mean.fish.size

fish.size.sd <- 10
sample.size <- 100
SE.mean.fish.size <- fish.size.sd/sqrt(sample.size)
SE.mean.fish.size
```

#Checkpoint 5: When we sample 30 fish, the standard error is 1.826, when we sample 50 fish, the standard error is 1.414, and when we sample 100 fish, the standard error is 1.

\############# NK Comments \##################

Looks good. If you wanted to do this all in one line, you would do the following.

`sample.size <- c(30,50,100)`

`fish.size.sd/sqrt(sample.size)`

\############# NK Comments \##################

```{r}
sample.sizes <- seq(from = 1, to = 100, by = 1)
sd.fish.length <- 5
SE.mean <- sd.fish.length/sqrt(sample.sizes)
plot(sample.sizes, SE.mean, typ = 'o', pch = 19,
xlab = 'Number of Fish In Sample', ylab = 'Standard Error of the Mean', ylim = c(0,5))
abline(h = 0, lty = 2)
```

```{r}
str(diabetes.df)
hist(diabetes.df$chol, xlab = 'Cholesterol')
```

```{r}
summary(diabetes.df$chol)

new.diabetes.df <- subset(diabetes.df, subset = !(is.na(chol)))
summary(new.diabetes.df$chol)
```

#Checkpoint 6: This subset function chooses all rows in the set where the chol colomn doesnt have NA values, so it filters out the rows with NA in the chol colomn.

\############# NK Comments \##################

Yep

\############# NK Comments \##################

```{r}
chol.mean <- mean(new.diabetes.df$chol)
chol.n <- length(new.diabetes.df$chol)
(chol.se <- sd(diabetes.df$chol, na.rm = T)/sqrt(chol.n))

possible.mean.chol <- seq(from = 150, to = 250, length = 10000)
poss.mean.prob <- dnorm(possible.mean.chol, mean = chol.mean, sd = chol.se)
plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220))

sampl.quantiles <- qnorm(c(0.025, 0.975), mean = chol.mean,  sd = chol.se) 

sampl.quantiles

plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220))
abline(v = sampl.quantiles, col = 'red', lty = 2)
abline(v = chol.mean, col = 'black', lty = 2, lwd = 3)
```

```{r}
chol.mean <- mean(new.diabetes.df$chol)
chol.n <- length(new.diabetes.df$chol)
(chol.se <- sd(diabetes.df$chol, na.rm = T)/sqrt(chol.n))

possible.mean.chol <- seq(from = 150, to = 250, length = 10000)
poss.mean.prob <- dnorm(possible.mean.chol, mean = chol.mean, sd = chol.se)
plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220))

sampl.quantiles <- qnorm(c(0.05, 0.95), mean = chol.mean,  sd = chol.se) 

sampl.quantiles

plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220))
abline(v = sampl.quantiles, col = 'red', lty = 2)
abline(v = chol.mean, col = 'black', lty = 2, lwd = 3)
```

#Checkpoint 7: #chol.mean \<- mean(new.diabetes.df$chol)
#chol.n <- length(new.diabetes.df$chol) \#(chol.se \<- sd(diabetes.df\$chol, na.rm = T)/sqrt(chol.n))

#possible.mean.chol \<- seq(from = 150, to = 250, length = 10000) #poss.mean.prob \<- dnorm(possible.mean.chol, mean = chol.mean, sd = chol.se) #plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220))

#sampl.quantiles \<- qnorm(c(0.05, 0.95), mean = chol.mean, sd = chol.se)

#sampl.quantiles

#plot(possible.mean.chol, poss.mean.prob, typ = 'l', xlab = 'Cholesterol Means', lwd = 3, ylab = 'Probability', xlim = c(195,220)) #abline(v = sampl.quantiles, col = 'red', lty = 2) #abline(v = chol.mean, col = 'black', lty = 2, lwd = 3)

\############# NK Comments \##################

Looks good

\############# NK Comments \##################

```{r}
our.model <- lm(chol ~ 1, data = new.diabetes.df)
confint(our.model, level = 0.95)
```

#Checkpoint 8: There was no checkpoint 8 on the lab assignment so I am writting this in. It skips from 7 to 9.

\############# NK Comments \##################

Whoops!

\############# NK Comments \##################

#Checkpoint 9: The confidence interval for the 99% confidence level (202.1085 to 213.583) is higher than the 95% confidence interval (203.4879 to 212.2037).

\############# NK Comments \##################

I might use the word "larger" rather than "higher", but I take your meaning.

\############# NK Comments \##################

```{r}
our.model <- lm(chol ~ 1, data = new.diabetes.df)
confint(our.model, level = 0.99)
```

```{r}
height.mean <- mean(new.diabetes.df$height)
height.n <- length(new.diabetes.df$height)
(height.se <- sd(diabetes.df$height, na.rm = T)/sqrt(height.n))

possible.mean.height <- seq(from = 150, to = 250, length = 10000)

our.modelheight <- lm(height ~ 1, data = new.diabetes.df)
confint(our.modelheight, level = 0.9)

height.mean <- mean(new.diabetes.df$height)
height.n <- length(new.diabetes.df$height)
(height.se <- sd(diabetes.df$height, na.rm = T)/sqrt(height.n))

possible.mean.height <- seq(from = 150, to = 250, length = 10000)

our.modelheight <- lm(height ~ 1, data = new.diabetes.df)
confint(our.modelheight, level = 0.95)

height.mean <- mean(new.diabetes.df$height)
height.n <- length(new.diabetes.df$height)
(height.se <- sd(diabetes.df$height, na.rm = T)/sqrt(height.n))

possible.mean.height <- seq(from = 150, to = 250, length = 10000)

our.modelheight <- lm(height ~ 1, data = new.diabetes.df)
confint(our.modelheight, level = 0.99)
```

#Checkpoint 10: The confidence intervals are listed with their code!

#90 Percent Interval: (65.69057 to 66.33966) #height.mean \<- mean(new.diabetes.df$height)
#height.n <- length(new.diabetes.df$height) \#(height.se \<- sd(diabetes.df\$height, na.rm = T)/sqrt(height.n)) #possible.mean.height \<- seq(from = 150, to = 250, length = 10000) #our.modelheight \<- lm(height \~ 1, data = new.diabetes.df) #confint(our.modelheight, level = 0.9)

#95 Percent Interval: (65.62811 to 66.40211) #height.mean \<- mean(new.diabetes.df$height)
#height.n <- length(new.diabetes.df$height) \#(height.se \<- sd(diabetes.df\$height, na.rm = T)/sqrt(height.n)) #possible.mean.height \<- seq(from = 150, to = 250, length = 10000) #our.modelheight \<- lm(height \~ 1, data = new.diabetes.df) #confint(our.modelheight, level = 0.95)

#99 Percent Interval: (65.50561 to 66.52462) #height.mean \<- mean(new.diabetes.df$height)
#height.n <- length(new.diabetes.df$height) \#(height.se \<- sd(diabetes.df\$height, na.rm = T)/sqrt(height.n)) #possible.mean.height \<- seq(from = 150, to = 250, length = 10000) #our.modelheight \<- lm(height \~ 1, data = new.diabetes.df) #confint(our.modelheight, level = 0.99)

\############# NK Comments \##################

Nice work.

\############# NK Comments \##################
